---
title: "Gotta Catch Em All - Pokémon Report"
author: "Bukola Elliott"
date: "2025-04-29"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
.libPaths('/home/runner/R/library')
# Load libraries
library(tidyverse)
library(ggplot2)
library(knitr)
library(png)
library(grid)
library(gt)
library(plotly)
library(curl)

# Custom theme for ggplot
theme_pokemon <- function() {
  theme_minimal(base_family = "Arial") +
    theme(
      plot.title      = element_text(face = "bold", size = 18, color = "#E3350D"),
      plot.subtitle   = element_text(face = "italic", size = 12, color = "#3B4CCA"),
      axis.text       = element_text(size = 10, color = "#3B4CCA"),
      panel.grid.major = element_line(color = "grey85"),
      plot.background  = element_rect(fill = "#FFFDF8", color = NA)
    )
}
```

## Introduction

In this analysis, I explore the April 1, 2025 Pokémon dataset from **TidyTuesday** to examine how different Pokémon **types** compare across key stats, particularly Speed and Attack. The analysis highlights the fastest and strongest Pokémon in each type, featuring sprite images, engaging trivia, and polished tables for a fun and insightful exploration of the Pokémon universe.

Additionally, I've created an interactive Shiny app that allows you to explore these metrics dynamically: [Click to View the Shiny App](https://bukks.shinyapps.io/pokemon_shiny/) [»]

For the full interactive experience, you can check out the RPubs version of this analysis:\
[Click to view Interactive Pokémon Analysis on RPubs](https://rpubs.com/Bukks/pokemon-analysis) [»]

## Load Pokémon Data

```{r}
# Load TidyTuesday data
pokemon_data <- tidytuesdayR::tt_load(2025, week = 13)
pokemon_data <- pokemon_data$pokemon_df

# Clean column names
pokemon_data <- janitor::clean_names(pokemon_data)
```

## Data Cleaning and Preparation

```{r}
# 1. Compute total stats and Add Legendary flag
pokemon_df <- pokemon_data %>%
  mutate(
    total = hp + attack + defense + special_attack + special_defense + speed,
    legendary = ifelse(total > 600, "Legendary", "Not Legendary")
  )

# 3. Fix missing or broken image URLs:
fallback_img <- "https://raw.githubusercontent.com/PokeAPI/sprites/master/sprites/items/poke-ball.png"

pokemon_df <- pokemon_df %>%
  mutate(
    url_image = ifelse(
      is.na(url_image) | url_image == "" | !grepl("^https?://", url_image),
      fallback_img,
      url_image
    )
  )
```

## Summary of Pokémon by Type

```{r}
# Count the number of Pokémon by their primary type
type_summary <- pokemon_data %>%
  count(type_1, sort = TRUE)

kable(type_summary, caption = "Pokémon Count by Primary Type")
```

Water-types are the most common (126), followed by Normal (111) and Grass (84), while Flying-only Pokémon are the rarest (4).

## Step 1: Summary Stats by Type

```{r}
# Compute average stats per primary type
pokemon_summary <- pokemon_df %>%
  group_by(type_1) %>%
  summarise(
    avg_speed  = mean(speed, na.rm = TRUE),
    avg_attack = mean(attack, na.rm = TRUE),
    avg_height = mean(height, na.rm = TRUE),
    avg_weight = mean(weight, na.rm = TRUE),
    .groups = 'drop'
  )

pokemon_summary
```

Dragon-types boast the highest average Attack (~108.7), and Flying-types dominate in average highest Speed (~102.5), whereas Fairy-types sit at the bottom for both stats (~61 Attack, ~53.6 Speed) and Steel-types are the heaviest (~225.6 kg).

## Step 2: Identify Top Pokémon per Type

```{r}
# Using species_id to build sprite URLs
fastest_pokemon <- pokemon_df %>%
  filter(!is.na(speed)) %>%
  group_by(type_1) %>%
  slice_max(order_by = speed, n = 1, with_ties = FALSE) %>%
  select(type_1, pokemon, speed, url_image, species_id) %>%
  ungroup()
fastest_pokemon
```

Deoxys-Speed tops all types with a Speed of 180, Ninjask rules Bug-types at 160, and Electrode is the fastest Electric-type at 150.

## Step 3: Combine Summaries

```{r}
# Merge summary stats with fastest Pokémon info
pokemon_type_stats <- pokemon_summary %>%
  left_join(fastest_pokemon, by = "type_1") %>%
  rename(
    fastest    = pokemon,
    fastest_sp = speed,
    img        = url_image
  )
```

This combined table brings together each type's average Speed and the single fastest Pokémon, making it easy to compare overall trends against individual standouts.

## Fixing: Ensure All Images Are Valid

```{r}
library(curl)

# Define fallback
fallback_img <- "https://raw.githubusercontent.com/PokeAPI/sprites/master/sprites/items/poke-ball.png"

# Create zero-padded IDs
pokemon_type_stats <- pokemon_type_stats %>%
  mutate(
    sprite  = sprintf("%03d", species_id),
    sprite_url = paste0(
      "https://raw.githubusercontent.com/HybridShivam/Pokemon/master/assets/images/",
      sprite, ".png"
    )
  )

# Validate & fallback to Poké Ball
fallback <- "https://raw.githubusercontent.com/PokeAPI/sprites/master/sprites/items/poke-ball.png"
valid   <- map_lgl(pokemon_type_stats$sprite_url,
                  ~ tryCatch(curl_fetch_memory(.x)$status_code == 200,
                             error = function(e) FALSE))

pokemon_type_stats <- pokemon_type_stats %>%
  mutate(img = if_else(valid, sprite_url, fallback))
```

## Step 4: Plotting Average Speed by Type with Images

```{r}
library(ggplot2)
library(ggimage)

ggplot(pokemon_type_stats,
       aes(y = reorder(type_1, avg_speed), x = avg_speed)) +
  geom_col(width = 0.6, fill = "#FFCB05", color = "#3B4CCA") +
  geom_image(aes(image = img),
             by = "height", size = 0.08,
             nudge_x = max(pokemon_type_stats$avg_speed) * 0.02) +
  geom_text(aes(label = fastest),
            hjust   = 0,
            nudge_x = max(pokemon_type_stats$avg_speed) * 0.07,
            size     = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.3))) +
  coord_cartesian(clip = "off") +
  labs(
    title    = "Pokémon Types by Average Speed",
    subtitle = "Sprites: Fastest Pokémon per Type",
    x        = "Average Speed",
    y        = "Type"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title     = element_text(face = "bold", size = 20),
    plot.subtitle  = element_text(size = 16, color = "#555555"),
    axis.title     = element_text(face = "bold", size = 14),
    axis.text      = element_text(size = 12),
    plot.margin    = margin(20, 50, 20, 20)
  )
```

## Step 5: Average Attack by Type

```{r}
# Identify strongest attackers
strongest_attackers <- pokemon_data %>%
  group_by(type_1) %>%
  filter(attack == max(attack, na.rm = TRUE)) %>%
  slice(1) %>%
  ungroup() %>%
  select(type_1, species_id, strongest = pokemon)

# Join the strongest attackers to the pokemon_type_stats
pokemon_type_stats <- pokemon_type_stats %>%
  left_join(strongest_attackers, by = "type_1") %>%
  mutate(
    img = paste0(
      "https://raw.githubusercontent.com/PokeAPI/sprites/master/sprites/pokemon/other/official-artwork/",
      species_id.y, ".png"
    ),
    strongest = ifelse(is.na(strongest.y), "Unknown", strongest.y)
  ) %>%
  select(-contains(".x"))

# Plot
ggplot(pokemon_type_stats, aes(y = reorder(type_1, avg_attack), x = avg_attack)) +
  geom_col(width = 0.6, fill = "#FF7F50", color = "#3B4CCA") +
  geom_image(aes(image = img), by = "height", size = 0.08, nudge_x = max(pokemon_type_stats$avg_attack) * 0.02) +
  geom_text(aes(label = strongest), hjust = 0, nudge_x = max(pokemon_type_stats$avg_attack) * 0.07, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.3))) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Pokémon Types by Attack",
    subtitle = "Sprites: Strongest Pokémon per Attack & Type",
    x = "Average Attack",
    y = "Type"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 20),
    plot.subtitle = element_text(size = 16, color = "#555555"),
    axis.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 12),
    plot.margin = margin(20, 50, 20, 20)
  )
```

## Top 10 Pokémon by Total Stats

```{r}
pokemon_data <- pokemon_data %>%
  mutate(total = hp + attack + defense + special_attack + special_defense + speed)

top_10 <- pokemon_data %>%
  arrange(desc(total)) %>%
  select(pokemon, type_1, type_2, total) %>%
  head(10)

kable(top_10, caption = "Top 10 Pokémon by Total Stats")
```

## Legendary Pokémon by Generation

```{r}
pokemon_data <- pokemon_data %>%
  mutate(legendary = ifelse(total > 600, "Legendary", "Not Legendary"))

legendary_summary <- pokemon_data %>%
  group_by(generation_id, legendary) %>%
  summarize(count = n()) %>%
  ungroup()

ggplot(legendary_summary, aes(x = factor(generation_id), y = count, fill = legendary)) +
  geom_col(position = "dodge") +
  labs(title = "Legendary Pokémon by Generation",
       x = "Generation", y = "Count")
```

```{r}
img_df <- pokemon_type_stats %>% select(type_1, fastest, img)

ggplot(img_df, aes(x = 1, y = type_1)) +
  geom_image(aes(image = img), size = 0.1, by = "height", asp = 1) +
  geom_text(aes(label = fastest),
            nudge_x = 0.12,
            hjust   = 0,
            size    = 3) +
  scale_y_discrete(
    limits = rev(unique(img_df$type_1)),
    expand = expansion(add = c(1, 1))
  ) +
  scale_x_continuous(limits = c(0.8, 1.6), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_void() +
  theme(
    plot.title  = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.margin = margin(20, 20, 20, 20)
  ) +
  labs(title = "Fastest Pokémon per Type: Image & Name")
```

## Conclusion

- **Speed vs. Attack:** Flying and Electric types lead in speed, Dragon and Fighting types in attack.
- **Stat Leaders:** Mega Mewtwo and Rayquaza top overall stats (>780), Primal Kyogre/Groudon next.
- **Generation Trends:** Generation 4 introduced the most Legendaries; later gens taper off.

*Thank you for exploring Pokémon stats with me!*
